Type-Safe Modular Parsing

### 2017-06-08 SLE_17/sections/CaseStudy.tex
revision

To demonstrate the utility of our modular parsing approach, we
implemented parsers of the first 18 calculi from book \textit{Types and Programming Languages} (TAPL) \cite{pierce2002types}. We compared our implementation with a non-modular implementation
we found online, which is also written in Scala and uses the same Packrat parsing library.
We counted source lines of code (SLOC) and measured execution time for both implementations.
The result suggests that our implementation saves 69\% code comparing with that non-modular one.

@@

We compared our implementation (named \ourimpl{}) with an implementation
available online\footnote{https://github.com/ilya-klyuchnikov/tapl-scala/} (named \ilyaimpl{}).
\ilyaimpl{} is suitable for comparison, because it is also
written in Scala using the same parser combinator library.
Furthermore, it includes parsers of all the 18 calculi we have, but
written in a non-modular way. Thus it is not able to reuse existing
code when those calculi share common features.

@@

\paragraph{Comparison Results}
Table \ref{tab:comparison} shows results of the comparison.
Let's only check \ourimpl{} and \ilyaimpl{} for now.
The overall result is that 69.2\% of code is reduced using our
approach, and our implementation is 42.7\% slower.

The good SLOC result is because of that the code of common language features
are reused lots of times in the whole case study. We can see that in the first two calculi
\lstinline{Arith} and \lstinline{Untyped} we are not better than \ilyaimpl{},
because in such two cases we do not reuse anything.
However in the following 16 calculi, we indeed reuse language components.
In particular, the calculi \inlinecode{EquiRec} and some others are only 22 lines
in our implementation, because we only compose existing codes.

@@

The right part of table \ref{tab:comparison} suggests that the difference of running time between
using Object Algebras and class is little, roughly 1\%.
The usage of longest match combinator slows the performance by 7\%. The main reason of slower
execution time may be the overall structure of the modular parsing approach, because we indeed have
more intermediate function calls and method overriding. However, it is worth mentioning that
because of the memoization technique of Packrat parsers, we are only constant times
slower, the algorithmic complexity is still the same.

>>>>

To demonstrate the utility of our modular parsing approach, we
implemented parsers for the first 18 calculi~\footnote{There are some more calculi in the book,
  but they are either not ported by the implementation we compare with, or just repeats the syntax of former ones.}
from the \textit{Types and Programming Languages} (TAPL)
\cite{pierce2002types} book.
We compared our implementation with a non-modular implementation
available online, which is also written in Scala and uses the same Packrat parsing library.
We counted source lines of code (SLOC) and measured execution time for both implementations.
The result suggests that our implementation saves 69\% code comparing
with that non-modular one, but there is a 43\% slowdown due to code modularity.

@@

We compared our implementation (named \ourimpl{}) with an implementation
available online\footnote{https://github.com/ilya-klyuchnikov/tapl-scala/} (named \ilyaimpl{}).
\ilyaimpl{} is suitable for comparison, because it is also
written in Scala using the same parser combinator library.
\ilyaimpl{} implements parsers 18 calculi in TAPL in a non-modular
way. Thus \ilyaimpl{} is not able to reuse existing
code when those calculi share common features.
\ourimpl{} implements the same 18 calculi, but reuse 
is possible due to modularity.

@@

\paragraph{Comparison Results}
Table \ref{tab:comparison} shows results of the comparison.
Let us only check \ourimpl{} and \ilyaimpl{} for now.
The overall result is that 69.2\% of code is reduced using our
approach, and our implementation is 42.7\% slower.

The good SLOC result is because of that the code of common language features
are reused many times in the whole case study. We can see that in the first two calculi
\lstinline{Arith} and \lstinline{Untyped} we are not better than \ilyaimpl{},
because in such two cases we do not reuse anything.
However in the following 16 calculi, we indeed reuse language components.
In particular, the calculi \inlinecode{EquiRec} and some others are only 22 lines
in our implementation, because we only compose existing code.

@@

The right part of table \ref{tab:comparison} suggests that the difference of running time between
using Object Algebras and class is little, roughly 1\%.
The usage of longest match combinator slows the performance by 7\%. The main reason of slower
execution time may be the overall structure of the modular parsing approach, because we indeed have
more intermediate function calls and method overriding. However, it is worth mentioning that
because of the memoization technique of Packrat parsers, we are only constant times
slower, the algorithmic complexity is still the same.
Since the slowdown seems to be caused by extra method
dispatching, in future work we wish to investigate techniques such as 
partial evaluation or meta-programming to eliminate such cost. The
work by B{\'e}guet and Manohar~\cite{Beguet:2014} is an interesting starting point.

### 2017-06-08 SLE_17/sections/ObjectAlgebras.tex
revision

\noindent Then all classes representing language constructs could be
extended to implement the operation. A first well-known problem is that such
approach is problematic in terms of type-safety (but see recent work
by Wang and Oliveira~\cite{wang2016expression}, which shows a technique that is type-safe
in many cases). More importantly, a second problem is that
even if that approach would work, the parsing code in \lstinline{VarParser} is no longer reusable!
The types \inlinecode{Expr, Lit, Add}, and so on, are all old types without the free variables operation.
To match the new ASTs, we have to substitute \inlinecode{NewExpr} for \inlinecode{Expr} (the
same for \inlinecode{Lit}, \inlinecode{Add}, ...). It requires either modification on existing code or casts.
The goal of semantic modularity motivates us to find a different approach for building ASTs.

@@

\lstinputlisting[linerange=8-17]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser looks fine, but it is not extensible. For example, we have demonstrated in Section~\ref{subsec:overriding} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, trying to do a similar method overriding for \inlinecode{pExpr} would require a type \inlinecode{VarAlg[E] =>} \inlinecode{Parser[E]}, which is a \emph{supertype} of the old type \inlinecode{Alg[E] =>} \inlinecode{Parser[E]}, since
the \emph{extended} Object Algebra interface appears in \emph{contravariant} position. This violates
overriding in Scala.

>>>>

\noindent Then all classes representing language constructs could be
extended to implement the operation. A first well-known problem is that such
approach is problematic in terms of type-safety (but see recent work
by Wang and Oliveira~\cite{wang2016expression}, which shows a technique that is type-safe
in many cases). More importantly, a second problem is that
even if that approach would work, the parsing code in \lstinline{VarParser} is no longer reusable!
The types \inlinecode{Expr, Lit, Add}, and so on, are all old types without the free variables operation.
To match the new ASTs, we have to substitute \inlinecode{NewExpr} for \inlinecode{Expr} (the
same for \inlinecode{Lit}, \inlinecode{Add}, ...). This requires either code modification or type casts.
The goal of semantic modularity motivates us to find a different approach for building ASTs.

@@

\lstinputlisting[linerange=8-17]{code/src/papercode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER_BAD
Such a parser looks fine, but it is not extensible. For example, we have demonstrated in Section~\ref{sec:inheritance} that method overriding is essential to update \inlinecode{pExpr} for an extended syntax. However, trying to do a similar method overriding for \inlinecode{pExpr} would require a type \inlinecode{VarAlg[E] =>} \inlinecode{Parser[E]}, which is a \emph{supertype} of the old type \inlinecode{Alg[E] =>} \inlinecode{Parser[E]}, since
the \emph{extended} Object Algebra interface appears in \emph{contravariant} position. This violates
overriding in Scala.
